---
title: "Jester - Collaborative Filtering"
author: "Jun Yan"
date: "November 7, 2017"
output: 
  html_document:
    toc: yes
---

## Starting Up

The ppt slides explaining the background behind the data set and collaborative filtering methods can be found here:

https://raw.githubusercontent.com/Tyllis/Data607/master/jester-collaborative-filtering.pptx

Load the packages and the file into R.

```{r message=F, warning=F}
library(dplyr)
library(knitr)
theURL <- "https://raw.githubusercontent.com/Tyllis/Data607/master/jester-data-1.csv"
jester <- read.csv(theURL, header = F, stringsAsFactors = F)
```

## Data Preparation

Save the 1st column in `num_rate`, then removed it from the data.

```{r}
# Save 1st column
num_rate <- jester[,1]   
# Remove 1st column
jester <- jester[,-1]
```

Inspect the data:

```{r}
# Check dimension
dim(jester)
num_row <- dim(jester)[1]
num_col <- dim(jester)[2]
# Rename rows and columns
names(jester) <- c(1:num_col)
row.names(jester) <- c(1:num_row)
# Show portion of data
kable(jester[1:10,1:15])
```

Replace all the value "99" by "NA"

```{r}
# Replace all value 99 by NA
jester <- as.data.frame(apply(jester, 2, function(x) replace(x, x==99, NA)))
```

Save the indices of NA. The result is a list containing `r num_row` row elements, each element contains the column indices of NA.

```{r}
# Save indices of NA
na_idx <- apply(jester, 1, function(x) which(is.na(x)))
```

## Create Similarity Matrix

Next, create several helper functions.

```{r}
# This function takes a vector and returns the magnitude of the vector.
mag <- function(x) sqrt(sum(x^2, na.rm=T))

# This function takes two vectors and returns the dot product of the two vectors.
dot <- function(x,y) sum(x*y, na.rm=T)

# This function takes two vectors and returns the cosine of the angles between the two vectors
cosine <- function(x,y) dot(x,y)/ (mag(x)*mag(y))

# Function that takes a vector and subtract the mean from each vector element.
center <- function(x) x-mean(x, na.rm=T)

# Function that takes two vectors and return the similarity of the vectors.
sim <- function(x,y) cosine(center(x), center(y))
```

Using the above helper functions, we can create the similarity matrix.

```{r}
# Create similarity matrix
sim_mat <- c()
for (col in 1:num_col){
  sim_mat <- rbind(sim_mat, apply(jester, 2, sim, jester[,col]))
}
```

The result is a matrix `sim_mat`. Each row contains the similarities vector. 

For example, the similarity between Joke #19 and Joke #66 is stored in row 19 col 66.

```{r}
# Check sim_mat[19,66]
sim(jester[,19], jester[,66]) == sim_mat[19,66]
```

Below code picks a pair of random jokes and check if the `sim_mat` stores the similarity value correctly.

```{r}
# Check a pair of random jokes
random_jokes <- sample(1:num_col, size=2)
joke1 <- random_jokes[1]
joke2 <- random_jokes[2]
print(paste(joke1, joke2))
sim(jester[,joke1], jester[,joke2])
sim_mat[joke1,joke2]
```

Also, another check is that the similarity matrix should be symmetric, meaning that sim_mat[i,j] = sim_mat[j,i]

```{r}
# Check symetric
sim_mat[joke1, joke2] == sim_mat[joke2, joke1]
```

Lastly, the diagonal of the similarity matrix should be all ones, because the similarity between each joke and itself should be 1. 

```{r}
# Check diagnal of similiarity matrix
diag(sim_mat)
```

Check a portion of the matrix.

```{r}
# Inspect the similarity matrix
dim(sim_mat)
kable(sim_mat[1:10,1:15], digits = 3)
```

## Neighborhood and Weighted Mean

Create a function that can search for the N-neighbors for a selected joke. The function takes three values:

- x = a vector containing the similarity values. This vector can be retrieved from `sim_mat` by calling `sim_mat[jokeID, ]`. 
- i = a vector containing the indices to exclude from the search. This vector is the indices of the jokes the user never rate before. This vector can be retrieved from `na_idx` created earlier, by calling `na_idx[[userID]]`. 
- n = a integer indicating how many neighbors to look for.

```{r}
# Function that takes a similarity vector x and a integer n and returns the top n neighbors but excluding the indices in the vector i. It returns a vector containing the similarity values, and the names of the vector are their indices.
n_neighbor <- function(x, i, n){
  # If na_idx is not empty, remove the indices from vector x
  if(length(na_idx[[i]])!=0) x=x[-na_idx[[i]]]
  # Sort x indecreasing order, take the top n+1, but remove the first one becaues the first one is always 1. 
  head(sort(x, decreasing = T), n+1)[-1]
} 
```

Create a function that calculates the weighted mean. The function will perform the following task:

- Given the userID, jokeID, and n, look for the n-neighbors of the joke relevant to the user
- Retrieve the similarity values of the n-neighbors of the joke
- Retrieve the rating values of the n-neighbors of the joke that the user had rated
- Calculate the weighted mean.

```{r}
# Function that takes a jokeID and a userID and estimate a rating based on collaborating filtering
cal_rating <- function(userID, jokeID, n){
  # look for n-neighbors
  neighbors <- n_neighbor(sim_mat[jokeID,], userID, n)  
  # retrieve the column indices of the neighbors
  idx <- as.integer(names(neighbors))
  # retrieve the rating values of the joke neighbors
  val <- jester[userID, idx]
  # calculate weighted mean
  weighted.mean(val, neighbors)
}
```

## Execution

Now it is ready to fill in the "NA" cells in the `jester` table. A nested for-loop implementation is used here, although a vectorized implementation may be possible. I also saved the predicted ratings separately in a list `predict_ratings`.

```{r}
jester_cf <- jester
predict_ratings <- list()
for (i in 1:num_row){
  temp <- c()
  if (length(na_idx[[i]])!=0){ 
    for (j in na_idx[[i]]){
      val <- cal_rating(i,j,5)
      jester_cf[i,j] <- val
      temp <- c(temp, val)
    }
    names(temp) <- na_idx[[i]]
  }
  predict_ratings[[i]] <- temp
}
```

Let's see portion of the data.

```{r}
# Inspect portion of data
kable(jester_cf[1:10,1:15], digits = 2)
```

The complete result can be found here:

https://raw.githubusercontent.com/Tyllis/Data607/master/jester_cf.csv
